---
title: "Processing Cobb County Tax Assessments Datasets"
author: "Melissa Juarez"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: yes
---

## Introduction

This analysis processes the tax assessment data for Cobb County. These datasets contain crucial digest and exact assessment data -- like parcel id, address, FMV, owner 1 &2, address, total tax, appeal value, mill values, exemptions, owner & site address in detail -- for residential and commercial properties together. This is super helpful in reducing inconsistencies and duplication when trying to merge residential and commercial data sets that are differently structured.

```{r setup, include = F}
knitr::opts_chunk$set(
	echo = FALSE,
	error = F,
	message = FALSE,
	warning = FALSE
)
 
options(scipen=999)

## load libraries
library(tidyverse)
library(dplyr)
library(here)
library(janitor)
library(readxl)
library(tidyr)
library(readr)
library(stringr)
library(writexl)
library(data.table)
```

## Digest Data
### Read and Join all Tax Data files

We will use the associated dictionaries for each file to set the data types for each column in the tax data.

```{r load-data, eval=FALSE, include=FALSE}

tax_paths <- list.files(here::here("output/cobb"), 
                           pattern = "TX502", 
                           recursive = T, full.names = T)

## if .txt then it is the data, if .xlsx then its the dictionary
tax_files_df <- data.frame(
  year = str_extract(tax_paths, "\\d{4}"),
  dict_path = ifelse(str_detect(tax_paths, "Layout"), tax_paths, NA),
  data_path = ifelse(str_detect(tax_paths, "PROD"), tax_paths, NA)
) %>%
  group_by(year) %>%
  summarise(
    dict_path = first(na.omit(dict_path)),
    data_path = first(na.omit(data_path)),
    .groups = "drop"
  )

## drop where either dict or data is not available
tax_files_df <- tax_files_df %>%
  filter(!is.na(data_path) & !is.na(dict_path))


## read in and clean data dictionaries
tax_files_df <- tax_files_df %>%
  mutate(
    data = map2(dict_path, data_path, ~ {
      dict <- read_csv(.x, show_col_types = FALSE)
      col_types_str <- paste(dict$type, collapse = "")
      read_csv(.y, col_types = col_types_str)
    })
  )

### @mel 5/15 -- need to add in 2021 and deal with diff column names....
tax_2021 <- read_csv(here("data/cobb/TAXDATA2021/TAXBILLS2021/TAXBILL_EXTRACT_2021_updated.csv"))
tax_2021 <- data.frame(year = 2021, data = tax_2021)

```

```{r see-common-cols}
## get col names from each file
column_names_list <- tax_files_df %>%
  mutate(cols = map(data, names)) %>%
  select(year, cols)

## get diff, similar column names
common_columns <- Reduce(intersect, column_names_list$cols)
all_columns <- Reduce(union, column_names_list$cols)
different_columns <- setdiff(all_columns, common_columns)

## create a table to view in which files certain columns differ 
column_presence <- map_dfr(different_columns, function(col) {
  tibble(
    column = col,
    year = column_names_list$year,
    present = map_lgl(column_names_list$cols, ~ col %in% .)
  )
}) %>%
  pivot_wider(names_from = year, values_from = present, values_fill = list(present = FALSE)) %>%
  arrange(column)

column_presence
```
problem w six flag special district SFEXM in 2015. two columns names the same thing. they keep the int class one which is coded as SFEXM...32. delete SFEXM...69

```{r}
## fix 2015 SFEXM cols
tax_files_df <- tax_files_df %>%
  mutate(data = map2(data, year, ~ if (str_detect(.y, "2015")) {
    .x %>%
      select(-`SFEXM...69`) %>%
      rename(
        SFEXM = `SFEXM...32`
      )
  } else {
    .x
  }))

tax_files_df <- tax_files_df %>%
  select(year, data) %>%  # Keep only necessary columns
  unnest(data) %>%
  rename(tax_year = year)

tax_files_df <- tax_files_df %>%
  select(-`(A “*” is used )`, `Blank space`) %>%
  clean_names()

## save data because it takes super long to read
write_csv(tax_files_df, file = here::here("output/cobb/RAW_cobb_digest_2011_2020.csv"))

```

### Cleaning: drop complete row duplicates

```{r}
raw_cobb_digest <- read.csv(here::here("output/cobb/RAW_cobb_digest_2011_2020.csv"))

cat("Total rows:", scales::comma(nrow(raw_cobb_digest)), 
    "\nUnique parcels:", scales::comma(length(unique(raw_cobb_digest$parid))))
```

```{r}
# drop complete dupes
init_len <- nrow(raw_cobb_digest)
cobb_digest <- distinct(raw_cobb_digest)
cat("Init len:", scales::comma(init_len), 
    "\nDropped dupes:", scales::comma(init_len - nrow(cobb_digest)))
```

### Eliminate duplicate TAXYR, PARID keys by selecting for greatest SQFT

These are parcels that contain multiple structures which are essentially ADUs; retain only the row with the largest living area square footage (main structure).

```{r}
init_len <- nrow(cobb_digest)

setDT(cobb_digest)
cobb_digest <- cobb_digest[
  order(tax_year, parid, -acres)  # sort by group + descending sfla
][
  , .SD[1], by = .(tax_year, parid)  # take top row per group
]

cat("Init len:", scales::comma(init_len), 
    "\nDropped parcels due to multiple entries in one year:", scales::comma(init_len - nrow(cobb_digest)))
```

### Check data consistency over the years

```{r}

count_of_records_parcel <- cobb_digest[, .(count_records = .N), by = parid]

count_missing_a_record <- sum(count_of_records_parcel$count_records < uniqueN(cobb_digest$taxyr))

tot_parcels_begin <- tot_parcels_begin <- merge(
  cobb_digest[taxyr == 2011],
  count_of_records_parcel,
  by = "parid",
  all.x = TRUE
)
begin_and_missing <- sum(tot_parcels_begin$count_records < 10)

pivoted_data <- dcast(
  cobb_digest[, .(parid, taxyr, record_exists = 1)],
  parid ~ taxyr,
  value.var = "record_exists",
  fill = 0
)

cat("Total number of unique parcels in DIGEST files:", length(unique(cobb_digest$parid)), 
    "\nNum parcels without a record for every year in period:", count_missing_a_record,
    "\nTotal number of parcels around at start of study period:", pivoted_data %>% filter(!is.na(`2011`)) %>% nrow(),
    "\nNum of parcels around at start that don't have an entry for every year:", begin_and_missing
    )

```
### Save cleaned data to excel
```{r}
cat("There is only one row per (taxyr, parid) group:", cobb_digest[, .N, by = .(taxyr, parid)][, max(N) == 1])
path <- paste0("output/cobb/DIGEST_cobb_2011_2020_", tolower(format(Sys.time(), "%b%d%Y")), ".csv")
write_csv(cobb_digest, file = here::here(path))

rm(combined_data, count_of_records_parcel, pivoted_data, tot_parcels_begin, begin_and_missing, count_missing_a_record, init_len)

```


## Appeal Data

```{r}
cobb_appeals <- read_excel(here("data/cobb/2012to2022CobbAppealData_All Divisions.XLSX"), skip = 1) %>%
  clean_names() %>%
  arrange(parid, tax_year) 

path <- paste0("output/cobb/APPEALS_cobb_2012_2022_", tolower(format(Sys.time(), "%b%d%Y")), ".csv")
write_csv(cobb_appeals, here(path))
```


