---
title: "Binding Paulding Datasets"
author: "Melissa Juarez"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: yes
---

## Introduction

This document takes the converted paulding data and binds them together so that data from all years is contained in a single file. As a reminder, 2011-2013 files are formatted in one way (one file per year) and the rest of the years have relational tables that may require merges to get the sales, appeals, and digest data. 

```{r}
library(dplyr)
library(tidyverse)
```


## Tyler 2011-2013

```{r get-tyler-data}
source <- here::here("output/paulding")
pattern <- paste(tyler_years, collapse = "|")
tyler_years <- 2011:2013

tyler_paths <- list.files(source, pattern = pattern, full.names = TRUE, recursive = TRUE)

nested_data <- tibble(file = tyler_paths) %>%
  mutate(data = map(file, ~ read_csv(.x, col_types = cols(.default = "c")) %>% janitor::clean_names()),
         tax_year = tyler_years)
```


```{r assess-tyler-column-names}
## get col names from each file
column_names_list <- nested_data %>%
  mutate(cols = map(data, names)) %>%
  select(tax_year, cols)

## get diff, similar column names
common_columns <- Reduce(intersect, column_names_list$cols)
all_columns <- Reduce(union, column_names_list$cols)
different_columns <- setdiff(all_columns, common_columns)

## create a table to view in which files certain columns differ 
column_presence <- map_dfr(different_columns, function(col) {
  tibble(
    column = col,
    tax_year = column_names_list$tax_year,
    present = map_lgl(column_names_list$cols, ~ col %in% .)
  )
}) %>%
  pivot_wider(names_from = tax_year, values_from = present, values_fill = list(present = FALSE)) %>%
  arrange(column)

column_presence
```

need to merge owner names and rename the LUC codes
owner_first_name cols are NA and entire names are contained in owner_last_name

```{r fix-tyler-column-names}
## fix 2012 owner cols
nested_data <- nested_data %>%
  mutate(data = map2(data, file, ~ if (str_detect(.y, "2012")) {
    .x %>%
      rename(
        owner_name_number_1 = owner_last_name_number_1,
        owner_name_number_2 = owner_last_name_number_2,
        owner_name_number_3 = owner_last_name_number_3,
        primary_land_use_code_luc_4th = primary_land_use_code_luc,
        subdivision_lot = subdivision_lot_number
      ) %>%
      select(-starts_with("owner_first_name_number"), -starts_with("owner_last_name_number"))
  } else {
    .x
  }))

```
now bind. but first we need to check data types to make sure they merge correctly

```{r bind-tyler-rows}
## convert all columns to character to prevent type mismatch issues before binding
nested_data <- nested_data %>%
  mutate(data = map(data, ~ mutate(.x, across(everything(), as.character))))

## row-bind all datasets, retaining tax_year col
long_table <- nested_data %>%
  select(tax_year, data) %>%  # Keep only necessary columns
  unnest(data)  # Flatten the nested data into one table

## convert back to appropriate column types
long_table <- type_convert(long_table, col_types = cols())

head(long_table)

## write file
path <- paste0("output/paulding/DIGEST_paulding_tyler_2011_2013_", tolower(format(Sys.time(), "%b%d%Y")), ".csv")
write_csv(long_table, here::here(path))

rm(col_names, column_names_list, column_presence, long_table, nested_data, all_columns, common_columns,
   different_columns, path, pattern, source, tyler_paths, tyler_years, year_folders)

```

## WinGap 2014-2022

For Wingap data, we need to read in:
-   PE_REALPROP.csv
-   PE_SALEINFO.csv
-   PE_OWNER.csv
-   PE_APPEALS.csv

realized that the data in appeals & sales files are cumulative so we only need to keep the most recent year

```{r}

source <- here::here("output/paulding")

wingap_paths <- list.files(source, full.names = TRUE, recursive = TRUE)
filtered_files <- wingap_paths[str_detect(wingap_paths, "20(1[4-9]|2[0-2]).*(PE_REALPROP|PE_SALEINFO|PE_OWNER|PE_APPEALS)")]


nested_data <- tibble(path = filtered_files) %>%
  mutate(tax_year = str_extract(path, "\\d{4}"),
         file = str_extract(path,"(?<=PE_)[^/.]+(?=\\.csv)"),
         data = map(path, ~ read_csv(.x) %>% janitor::clean_names()))

## write appeals data
all_appeals <- nested_data %>% 
  filter(file == "APPEALS" & tax_year == max(tax_year)) %>%
  select(data) %>%
  unnest()


path <- paste0("output/paulding/APPEALS_paulding_1999_2022_", tolower(format(Sys.time(), "%b%d%Y")), ".csv")
write_csv(all_appeals, here::here(path))

rm(all_appeals)

## write sales data
all_sales <- nested_data %>% 
  filter(file == "SALEINFO" & tax_year == max(tax_year)) %>%
  select(data) %>%
  unnest()

path <- paste0("output/paulding/SALES_paulding_1886_2022_", tolower(format(Sys.time(), "%b%d%Y")), ".csv")
write_csv(all_sales, here::here(path))

rm(all_sales)

```


### Owner data

```{r}
owner_data <- nested_data %>% 
  filter(file  == "OWNER")

## get col names for each file
column_names_list <- owner_data %>%
  mutate(cols = map(data, names)) %>%
  select(tax_year, cols)

## get diff, similar column names
common_columns <- Reduce(intersect, column_names_list$cols)
all_columns <- Reduce(union, column_names_list$cols)
different_columns <- setdiff(all_columns, common_columns)

## create a table to view in which files certain columns differ 
column_presence <- map_dfr(different_columns, function(col) {
  tibble(
    column = col,
    tax_year = column_names_list$tax_year,
    present = map_lgl(column_names_list$cols, ~ col %in% .)
  )
}) %>%
  pivot_wider(names_from = tax_year, values_from = present, values_fill = list(present = FALSE)) %>%
  arrange(column)


column_presence

```

only two columns are added later, not a problem for binding.

```{r}
## convert zip to character to prevent type mismatch issues before binding
owner_data <- owner_data %>%
  mutate(data = map(data, ~ mutate(.x, zip = as.character(zip))))

## row-bind all datasets, retaining tax_year
owner_data <- owner_data %>%
  select(tax_year, data) %>%  # Keep only necessary columns
  unnest(data)  # Flatten the nested data into one table

owner_data <- owner_data %>%
  rename_with(~ paste0("owner_", .), .cols = -c(1,2))

## dedupe owner rows
owner_data <-owner_data %>%
  distinct()

head(owner_data) ## keep this to merge w digest later

```




```{r}
digest_data <- nested_data %>% 
  filter(file  ==  "REALPROP")

## get col names for each file
column_names_list <- digest_data %>%
  mutate(cols = map(data, names)) %>%
  select(tax_year, cols)

## get diff, similar column names
common_columns <- Reduce(intersect, column_names_list$cols)
all_columns <- Reduce(union, column_names_list$cols)
different_columns <- setdiff(all_columns, common_columns)

## create a table to view in which files certain columns differ 
column_presence <- map_dfr(different_columns, function(col) {
  tibble(
    column = col,
    tax_year = column_names_list$tax_year,
    present = map_lgl(column_names_list$cols, ~ col %in% .)
  )
}) %>%
  pivot_wider(names_from = tax_year, values_from = present, values_fill = list(present = FALSE)) %>%
  arrange(column)

sort(common_columns)
column_presence

```

It looks like only new columns are being added so we can bind just fine. I am only looking for if there are any columns that are renamed so that change them to have the same name between years.

```{r}
# convert unit column to character to prevent type mismatch issues before binding
digest_data <- digest_data %>%
  mutate(data = map(data, ~ mutate(.x, 
                                   unit = as.character(unit))))

# row-bind all datasets, retaining tax_year
long_table <- digest_data %>%
  select(tax_year, data) %>% 
  unnest(data) %>% 
  distinct() # dedupe

head(long_table)

```

merge owner and digest on ownkey and tax year
```{r}
digest <- long_table %>%
  merge(owner_data, by = c("tax_year", "ownkey"), all.x = T)


```

save the data
```{r}
path <- paste0("output/paulding/DIGEST_paulding_wingap_2014_2022_", tolower(format(Sys.time(), "%b%d%Y")), ".csv")
write_csv(digest, here::here(path))

```

